---
title: "US Census Demographics (2000–2020)"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)

# Load state boundaries and decennial census data
us_geo  <- read_rds("data/us_geo.rds")
us_race <- read_rds("data/us_race_2000_2020.rds")

# Reshape data to wide format to compute percentage point changes (2000 to 2020)
race_wide <- us_race |>
  mutate(pct = round(100 * (value / summary_value), 1)) |>
  select(GEOID, variable, year, pct) |>
  pivot_wider(names_from = c(variable, year), values_from = pct) |>
  mutate(
    diff_hisp  = round(hispanic_2020 - hispanic_2000, 1),
    diff_black = round(black_2020 - black_2000, 1),
    diff_white = round(white_2020 - white_2000, 1),
    diff_asian = round(asian_2020 - asian_2000, 1)
  )

# Combine with state geometries and reproject to WGS84
us_map <- left_join(us_geo, race_wide, by = "GEOID") |>
  st_transform(4326)

# Create hover labels summarizing all four groups for each state
labels <- sprintf(
  "<strong>%s</strong><br/>
   Hispanic: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   White: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   Black: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   Asian: %s%% &rarr; %s%% (%+0.1f pts)",
  us_map$NAME,
  us_map$hispanic_2000, us_map$hispanic_2020, us_map$diff_hisp,
  us_map$white_2000, us_map$white_2020, us_map$diff_white,
  us_map$black_2000, us_map$black_2020, us_map$diff_black,
  us_map$asian_2000, us_map$asian_2020, us_map$diff_asian
) |> lapply(HTML)

# Define color scales for each demographic group
pal_hisp  <- colorNumeric("YlOrRd", domain = us_map$diff_hisp)
pal_white <- colorNumeric("Blues", domain = us_map$diff_white)
pal_black <- colorNumeric("Purples", domain = us_map$diff_black)
pal_asian <- colorNumeric("Greens", domain = us_map$diff_asian)
```

## Two Decades of Demographic Change Across the US

Between 2000 and 2020, the United States saw significant demographic shifts across all major racial and ethnic groups:

- **Hispanic**: Substantial increases nationwide, particularly in southern and western states.
- **White (non-Hispanic)**: Decreased as an overall percentage of population in almost all states.
- **Black**: Steady population shares with shifts in southern and metropolitan states.
- **Asian**: Steady percentage growth across major metropolitan and coastal states.

Use the radio buttons in the upper right to toggle which demographic shift colors the map. Hover over any state to view the full 20-year shift for all four groups.

```{r}
#| label: interactive-map
#| message: false
#| echo: false

leaflet(us_map, width = "100%", height = "540px") |>
  addTiles() |>
  addPolygons(
    group = "Hispanic Change",
    fillColor = ~pal_hisp(diff_hisp),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "White Change",
    fillColor = ~pal_white(diff_white),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Black Change",
    fillColor = ~pal_black(diff_black),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Asian Change",
    fillColor = ~pal_asian(diff_asian),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addLayersControl(
    baseGroups = c("Hispanic Change", "White Change", "Black Change", "Asian Change"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
