---
title: "US & Florida Demographics (2000–2020)"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)
library(htmltools)

# Load state and Florida county data
us_geo   <- read_rds("data/us_geo.rds")
us_race  <- read_rds("data/us_race_2000_2020.rds")
fl_geo   <- read_rds("data/fl_geo.rds")
fl_race  <- read_rds("data/fl_race_2000_2020.rds")

# Helper to compute percentage point changes (2000 to 2020)
calc_wide <- \(df) {
  df |>
    mutate(pct = round(100 * (value / summary_value), 1)) |>
    select(GEOID, variable, year, pct) |>
    pivot_wider(names_from = c(variable, year), values_from = pct) |>
    mutate(
      diff_hisp  = round(hispanic_2020 - hispanic_2000, 1),
      diff_black = round(black_2020 - black_2000, 1),
      diff_white = round(white_2020 - white_2000, 1),
      diff_asian = round(asian_2020 - asian_2000, 1)
    )
}

us_wide <- calc_wide(us_race)
fl_wide <- calc_wide(fl_race)

# Keep other 49 states at state level, filter out Florida
us_other_states <- left_join(us_geo, us_wide, by = "GEOID") |>
  filter(NAME != "Florida") |>
  st_transform(4326)

# Keep Florida at county level
fl_counties <- left_join(fl_geo, fl_wide, by = "GEOID") |>
  st_transform(4326)

# Combine both into a single map dataset
combined_map <- bind_rows(us_other_states, fl_counties)

# Create hover labels summarizing all four groups
labels <- sprintf(
  "<strong>%s</strong><br/>
   Hispanic: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   White: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   Black: %s%% &rarr; %s%% (%+0.1f pts)<br/>
   Asian: %s%% &rarr; %s%% (%+0.1f pts)",
  combined_map$NAME,
  combined_map$hispanic_2000, combined_map$hispanic_2020, combined_map$diff_hisp,
  combined_map$white_2000, combined_map$white_2020, combined_map$diff_white,
  combined_map$black_2000, combined_map$black_2020, combined_map$diff_black,
  combined_map$asian_2000, combined_map$asian_2020, combined_map$diff_asian
) |> lapply(HTML)

# Define color scales for each demographic group
pal_hisp  <- colorNumeric("YlOrRd", domain = combined_map$diff_hisp)
pal_white <- colorNumeric("Blues", domain = combined_map$diff_white)
pal_black <- colorNumeric("Purples", domain = combined_map$diff_black)
pal_asian <- colorNumeric("Greens", domain = combined_map$diff_asian)
```

## Two Decades of Demographic Change Across the US and Florida

Between 2000 and 2020, the United States saw significant demographic shifts across all major racial and ethnic groups:

- **Hispanic**: Substantial increases nationwide, particularly across the Sun Belt and Florida counties.
- **White (non-Hispanic)**: Decreased as an overall percentage of population across nearly all states and counties.
- **Black**: Steady population shares with shifts across southern regions and metropolitan areas.
- **Asian**: Steady percentage growth across major metropolitan and suburban areas.

Use the radio buttons in the upper right to toggle which demographic shift colors the map. Hover over any state (or Florida county) to view the full 20-year shift for all four groups.

```{r}
#| label: interactive-map
#| message: false
#| echo: false

leaflet(combined_map, width = "100%", height = "540px") |>
  setView(lng = -98.5, lat = 39.8, zoom = 4) |>
  addTiles() |>
  addPolygons(
    group = "Hispanic Change",
    fillColor = ~pal_hisp(diff_hisp),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "White Change",
    fillColor = ~pal_white(diff_white),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Black Change",
    fillColor = ~pal_black(diff_black),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addPolygons(
    group = "Asian Change",
    fillColor = ~pal_asian(diff_asian),
    fillOpacity = 0.7, weight = 1, color = "white",
    label = labels
  ) |>
  addLegend(
    pal = pal_hisp, values = ~diff_hisp,
    group = "Hispanic Change", position = "bottomright",
    title = "Hispanic (pts)"
  ) |>
  addLegend(
    pal = pal_white, values = ~diff_white,
    group = "White Change", position = "bottomright",
    title = "White (pts)"
  ) |>
  addLegend(
    pal = pal_black, values = ~diff_black,
    group = "Black Change", position = "bottomright",
    title = "Black (pts)"
  ) |>
  addLegend(
    pal = pal_asian, values = ~diff_asian,
    group = "Asian Change", position = "bottomright",
    title = "Asian (pts)"
  ) |>
  addLayersControl(
    baseGroups = c("Hispanic Change", "White Change", "Black Change", "Asian Change"),
    options = layersControlOptions(collapsed = FALSE)
  )
```
